---
title: "Time Series - Homework 1"
author: "Matt Isaac"
date: "September 18, 2018"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## I. Data Analysis

1. The data set I chose is `nhtemp`, which contains data for the average yearly temperatures in New Haven from 1912 to 1971.  

* Plot the data  
```{r}
plot(nhtemp, xlab = "Year", ylab = "Temperature", 
     main = "Average Yearly Temperatures in New Haven")
```

* Identify any trend or seasonality or both in your data.  
I see a slight upward linear trend present in these data. I don't see any evidence of seasonality/period in these data. 

* Transform the data into a stationary process and plot the transformed data  
```{r}
nhtemp_st <- diff(nhtemp, differences = 1)
plot(nhtemp_st, xlab = "Year", ylab = "Temperature", 
     main = "Average Yearly Temperatures in New Haven (Transformed")
```

2. Consider the Nile and the JohnsonJohnson data in R Datasets package. For each one
of the two:  

* Transform the original time series into a stationary process.  
* Plot the original data, transformed data, and the sample ACF of the transformed
data on one page.

#### Nile data  
* Transform into stationary process.
* Plots  
```{r, fig.height=10}
Nile_st <- diff(Nile, differences = 1)
par(mfrow = c(3,1))
plot(Nile, main = "Yearly Flow of the River Nile", xlab = "Year", ylab = "Flow")
plot(Nile_st, main = "Yearly Flow of the River Nile (Transformed)", xlab = "Year", ylab = "Flow")
acf(Nile_st, main = "ACF")

```

#### JohnsonJohnson data  
* Transform into stationary process.  
* Plots  
```{r, fig.height=10}
par(mfrow = c(3,1))
plot(JohnsonJohnson, main = "Quarterly Sales for Johnson and Johnson", ylab = "sales")
lnJJ <- log(JohnsonJohnson)
lnJJ_st1 <- diff(lnJJ, diff = 2)
lnJJ_st2 <- diff(lnJJ_st1, lag = 4)
plot(lnJJ_st2, main = "Quarterly Sales for Johnson and Johnson (Transformed)", ylab = "sales")
acf(lnJJ_st2, main = "ACF")
```
